---
title: "Chapter 7"
author: "Zach Brown"
date: "July 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 7.1 Introduction
## 7.1.1 The Monte Carlo Method

If we want to approximate the expected value of the sum of 3 random exponential variables, we can replicate that process multiple times.  In this case, we performed it 6000 times.
```{r}
empirical = replicate(6000, sum(rexp(n = 3)))
mean(empirical)
sd(empirical)
```
The actual values for the mean and standard deviation are 3 and the square root of 3 respectively, so our approximations come very close to the real values.  

If we wanted to find the chance that the sum is at most 5, it is a little harder to calculate analytically, but still very easy to estimate.
```{r}
sum(empirical <= 5)/length(empirical)
```

We also might want to look at the impact of the rate parameter on the distribution.  We can select multiple rate values to use in our simulation.
```{r}
rates = c(seq(0.1, 1, by = 0.1), seq(2, 7, by = 1))

samples = lapply(rates, function(r) {
  replicate(6000, sum(rexp(n = 3, rate = r)))
})
```

# 7.2 Exploring the Random Process

We will now model the completion time (or lifetime) of a process using an exponential distribution.  The exponential distribution has a rate parameter kappa.  The density is kappa * exp^(-kappa * x), where x > 0.  

We'll start with a kappa of 0.3
```{r}
kappa = 0.3
d0 = rexp(1, rate = kappa)
d0
```
Our first job starts at time 0 and finishes at time 3.195.  Now we can simulate the spawning of child jobs.  We will use a rate of 0.5.
```{r}
lambda = 0.5
birth1 = rexp(1, rate = lambda)
birth1
```
We can generate the inter-arrival time between the first and second offspring.
```{r}
itime = c(birth1, rexp(1, rate = lambda))
itime
```
This means the second job arrives 1.486 seconds after the first.  
A third child process can also be generated
```{r}
itime = c(itime, rexp(1, rate = lambda))
itime
```
The third child process is generated 2.37 seconds later.  

We can convert the inter-arrival times to birth times by using a cumulative sum.  
```{r}
cumsum(itime)
```
The third time occurs after the parent job has finished, so we will only keep the first two child jobs.
```{r}
btime = cumsum(itime)
btime = btime[btime < d0]
```

Each job must wait for its parent job to finish running before it begins running.  So the lifecycle of a job can be split up into the time that it is waiting for its parent job to complete and the time that it is running.  The runtime for each child job has the same distribution as the runtime for the parent job.  
Sinc ethe child jobs cannot start running until the parent job completes, we can generate their start times by adding their birth times to the parent job's runtime
```{r}
dtime = d0 + rexp(n = length(btime), rate = kappa)
dtime
```
In this instance, the second born is the first to die and the first born is the last to die.

## 7.3 Generating Offspring
Now we can encapsulate the work done above in a function to make it easily repeatable.  We will denote birth time, completion time, rate for the inter-arrival of offspring, and a rate for the run times of the offspring with the parameters bTime, cTime, lambda, and kappa respectively.

```{r}
genKids = function(bTime, cTime, lambda = 0.5, kappa = 0.3) {
  #Parent job born at bTime and completes at cTime
  
  # Birth time of first child
  mostRecent = rexp(1, rate = lambda) + bTime
  kidBirths = numeric()
  
  while (mostRecent < cTime) {
    kidBirths = c(kidBirths, mostRecent)
    mostRecent = mostRecent + rexp(1, rate = lambda)
  }
  
  # Generate lifetimes for all offspring
  numKids = length(kidBirths)
  runtime = rexp(numKids, rate = kappa)
  kidCompletes = rep(cTime, numKids) + runtime
  
  data.frame(births = kidBirths,
             completes = kidCompletes)
}
```

Now we can test out this function
```{r}
genKids(1, 6)
```
```{r}
genKids(1, 6)
```
```{r}
genKids(1, 6)
```
This function could also be implemented using recursion
```{r}
genBirth = function(currentTime, cTime,
                    births = numeric(), lambda = 0.5) {
  
  # Generate birth time of next job after currentTime
  mostRecent = rexp(1, rate = lambda) + currentTime
  
  if (mostRecent > cTime)
    return(births)
  else {
    births = c(births, mostRecent)
    genBirth(currentTime = mostRecent, cTime, births, lambda)
  }
}
```

```{r}
genKidsR =
  function(bTime, cTime, lambda = 0.5, kappa = 0.3) {
    # Parent job born at bTime and completes at cTime
    
    kidBirths = genBirth(bTime, cTime, lambda = lambda)
    
    # Generate lifetimes for all offspring
    numKids = length(kidBirths)
    runtime = rexp(numKids, rate = kappa)
    kidDeaths = rep(cTime, numKids) + runtime
    
    data.frame(births = kidBirths,
               completes = kidDeaths)
  }
```

## 7.3.1 - Checking the Results
We can check that the function is working properly by running it many times and seeing if the average numberof children generated matches with the expected number.  We expect this number to be lambda * (deathtime - birthtime).  In this case, that works out to be 0.5 * (6 - 1) or 2.5.
```{r}
numKids = replicate(1000, nrow(genKids(1, 6)))
mean(numKids)
```
The standard deviation of the number of children is the square root of 2.5 (or 1.58), which means the standard error is the square root of 2.5/1000 (or 0.05).  Our result is within one standard error.  The distribution should follow the Poisson distribution.  we can check that as well.
```{r}
eprobs = table(numKids)/length(numKids)
probs = dpois(x = 0:max(numKids), lambda = 2.5)
```
```{r}
plot(eprobs, type = "h",
     ylab = "Proportion", xlab = "Number of Offspring")
segments(x0 = 0.1 + 0:max(numKids), y0 = rep(0, max(numKids)),
         y1 = probs, col = "grey", lwd = 2)
```
We can see that the numbers generated do, in fact, follow a Poisson distribution.  
## 7.3.2 - Consider Alternative Implementations
We can also implement this functonality in a different way, given some things we know about the Poisson distribution.  We know that the number of births follows the Poisson(lambda t) distribution where lambda is the rate per unit of time.  Also, if we know the number of events in the time interval, we also know that they follow the Uniform Distribution.  So instead of generating offspring one after the other using inter-arrival times, we can generate all of the offspring at once and then determine their birth times.
```{r}
genKidsU = 
  function(bTime, cTime, lambda = 0.5, kappa = 0.3) {
         # Generate the birth times and assassination times
         # for the children of a job who is born at bTime 
         # and completed at cTime.
    lambda = (cTime - bTime) * lambda
    numKids = rpois(1, lambda = lambda)
    kidBirths = sort(runif(numKids, min = bTime, max = cTime))
    
         # generate lifetimes for each offspring
    runtime = rexp(numKids, rate = kappa)
    kidDeaths = rep(cTime, numKids) + runtime
    
    return(data.frame(births = kidBirths, completes = kidDeaths))
  }
```

# 7.4 - Profiling and Improving Our Code
Now we can compare the runtime of each of the genKids function with the genKidsU function.
```{r}
seedx = 116201412
set.seed(seedx)

time1 = system.time( replicate(4000, genKids(1, cTime = 9)) )
time2 = system.time( replicate(4000, genKidsU(1, cTime = 9)) )

time1/time2
```
The approach using the while loop took about 89% of the time that the approach using the Poisson and Uniform properties to run.  We will see if this holds when the completion time of the parent is increased.
```{r}
time1 = system.time( replicate(4000, genKids(1, cTime = 100)) )
time2 = system.time( replicate(4000, genKidsU(1, cTime = 100)) )

time1/time2
```
When the parent completion time increased to 100, there is time for more child nodes to be created.  Now the while loop solution is about 77% slower.  

We can use the Rprof() function to investigate which pieces of the function are taking the longest to run.
```{r}
set.seed(seedx)
Rprof("profGenKids1.out")
invisible(replicate(1000, genKids(1, cTime = 100)))
Rprof(NULL)

summaryRprof("profGenKids1.out")$by.self
```

```{r}
set.seed(seedx)
Rprof("profGenKidsU.out")
invisible(replicate(1000, genKidsU(1, cTime = 100)))
Rprof(NULL)

summaryRprof("profGenKidsU.out")$by.self
```
We can see that rexp accounts for about 23% of the runtime in genKids, but only about 7% in genKidsU.

# 7.5 - From One Job's Offspring to an Entire Generation
Now we can modify the genKidsU function to create child processes for multiple generations
```{r}
genKidsV = function(bTimes, cTimes, lambda = 0.5, kappa = 0.3) {
  # bTimes & cTimes  - vector of birth and completion times
  
  # determine how many children each job has
  parentAge = cTimes - bTimes
  numKids = rpois(n = length(parentAge),
                  lambda = lambda*parentAge)
  
  # Determine the birth and completion times of the children
  mapply(function(n, min, max) {
    births = sort(runif(n, min, max))
    runtimes = rexp(n, rate = kappa)
    completes = rep(max, n) + runtimes
    data.frame(births, completes)
  },
  n = numKids, min = bTimes, max = cTimes, SIMPLIFY = FALSE)
}
```

# 7.6 - Unit Testing
Now we can set up some tests for this new branching function  

First, we will test 3 jobs with different birth and completion times.
```{r}
bTimes1 = 1:3
cTimes1 = c(3, 10, 15)

seed1 = 12062013
set.seed(seed1)

kids = genKidsV(bTimes1, cTimes1)
kids
```
This generates 3 processes with 1, 3, and 4 child processes respectively.
```{r}
kids2 = genKidsV(bTimes1, cTimes1)
sapply(kids2, nrow)
```
Since we did not reset the seed after the first run, we get different results when running the genKidsV function again with the same parameters.
```{r}
set.seed(seed1)
kids3 = genKidsV(bTimes = bTimes1, cTimes = cTimes1)
identical(kids, kids3)
```
We can see that if we reset the seed, the function once again generates the same results that it originally did.

# 7.7 - A Structure for the Function's Return Value
In order to keep track of which child processes were generated by which parent processes, we need to modify the genKidsV function to add an identifier for each process.
```{r}
genKidsV = function(bTimes, cTimes, parentID, lambda = 0.5, kappa = 0.3) {
       # Determine how many children each job has
  parentAge = cTimes - bTimes
  numKids = rpois(n = length(parentAge), lambda = lambda * parentAge)
  
  if (sum(numKids) == 0) return(NULL)
  
       # Determine the birth times of the children  
  kidStats = 
    mapply(function(n, min, max) {
      births = sort(runif(n, min, max))
      runtimes = rexp(n, rate = kappa)
      completes = rep(max, n) + runtimes
      data.frame(births, completes)
    },
           n = numKids , min = bTimes, max = cTimes, 
           SIMPLIFY = FALSE)
  
 
  return(data.frame(parentID = rep(parentID, numKids),
                    kidID = 1:sum(numKids), 
                    births = unlist(lapply(kidStats, "[[", "births")), 
                    completes = unlist(lapply(kidStats,"[[", "completes"))
                    ))
}
```

```{r}
set.seed(seed1)
genKidsV(bTimes1, cTimes1, parentID = letters[1:3])
```

# 7.8 - The Family Tree: Simulating the Branching Process
Now that we can create a generation using the genKidsV function, we can use this to model the entire process.  To guard against the process running for an infinite number of generations, we will pass a parameter to this new function to set the maximum number of generations to be modeled.
```{r}
familyTree = function(lambda = 0.5, kappa = 0.3, maxGen = 10) {
  # maxGen - maximum number of generations to observe
  # Return value - a list with 1 data frame per generation.
  allGens = vector(mode = "list", length = maxGen)
  
  # Generate the root of the tree
  allGens[[1]] = data.frame(parentID = NA, kidID = 1, births = 0,
                            completes = rexp(1, rate = kappa))
  
  # Generate future generations, one at a time.
  for (i in 2:maxGen) {
    nextGen = genKidsV(bTimes = allGens[[ (i - 1) ]]$births,
                       cTimes = allGens[[ (i - 1) ]]$completes,
                       parentID = allGens[[ (i - 1) ]]$kidID,
                       lambda = lambda, kappa = kappa)
    if (is.null(nextGen)) return(allGens[ 1:(i - 1) ])
    allGens[[ i ]] = nextGen
  }
  
  return(allGens)
}
```
Now we can test this function
```{r}
set.seed(seed1)
tree = familyTree(lambda = 0.4, kappa = 1, maxGen = 10)
```
Since there is so much data generated by the familyTree function, it is easier to investigate it by plotting it.
```{r}
set.seed(seed1)
g = familyTree(lambda = 0.4, kappa = 1, maxGen = 10)

maxLife = max(sapply(g, function(gen) max(gen$completes)))
numFamily = sum(sapply(g, nrow))

plot(0,0, ylim = c(0.5, numFamily + 0.5), xlim = c(0, maxLife),
     xlab = "Time", ylab = "", type ="n",
     axes = FALSE)

box()
axis(1)
numGen = length(g)
numKids = sapply(g, nrow)
treeN = g[ 2:(numGen + 1) ]
birthNum = c(0, cumsum(sapply(g, nrow))[ -length(g)])
axis(2, at = birthNum + 1, 
     labels = paste("Gen", 1:numGen), tick = FALSE, las = 1)

mapply(function(gen, nextGen, birthNum) {
  birthOrder = birthNum + (1:nrow(gen))
  segments(x0 = gen$births, x1 = gen$completes, y0 = birthOrder,
          lwd = 3, col = "grey")
  abline(h = 0.5 + max(birthOrder), lty = 2, col="black" )
  if (all(!is.na(nextGen$births)))
  points(x = nextGen$births, y = birthOrder[nextGen$parentID],
         pch = 4)
  },
  gen = g, nextGen = treeN, birthNum = birthNum )
```

Let's create a new tree with a different seed.
```{r}
seed2 = 12212013
set.seed(seed2)
tree = familyTree(lambda = 0.3, kappa = 0.5, maxGen = 10)
```
We can tell how many generations are in a tree by taking its length.
```{r}
length(tree)
```
We can also tell how many children are in each generation by counting the rows.
```{r}
sapply(tree, nrow)
```
Similarly, the total number of children can be obtained by summing these row counts.
```{r}
sum(sapply(tree, nrow))
```
As the number of generations increases, the number of children increases exponentially.  Therefore, we need a better method for analysis than the previous plot, as that does not scale to thousands of children.
```{r}
set.seed(seed2)
tree = familyTree(lambda = 1, kappa = 0.5, maxGen = 10)
```
```{r}
length(tree)
```
```{r}
sapply(tree, nrow)
```
```{r}
sum(sapply(tree, nrow))
```
This time, the tree produced over 670,000 children.  Let's run it again and see if this is normal.

```{r}
tree = familyTree(lambda = 1, kappa = 0.5, maxGen = 10)
sum(sapply(tree, nrow))
```
This time the tree has almost a million children.  Clearly this can get out of hand quickly and without limiting the parameters, the familyTree function might never return.  In order to find the best way to do that, we can first investigate the birth and completion times for each generation.
```{r}
sapply(tree, function(gen) range(gen$births))
```
```{r}
sapply(tree, function(gen) range(gen$completes))
```
We can see from these ranges that there are children in generation 10 that were born before some children in generation 4.  
```{r}
familyTreeT = function(lambda = 0.5, kappa = 0.3, maxTime = 8) {
       # maxTime - maximum length of time to observe the process
  allGens = list()
  
       # Generate the root of the 
  allGens[[1]] = data.frame(parentID = NA, kidID = 1, 
                            births = 0, 
                            completes = rexp(1, rate = kappa))
  
       # Generate the future generations, one at a time.
  numGens = 1
  while (TRUE) {
    nextGen = genKidsV(bTimes = allGens[[ numGens ]]$births,
                       cTimes = allGens[[ numGens ]]$completes,
                       parentID = allGens[[ numGens ]]$kidID,
                       lambda = lambda, kappa = kappa)
    
    if ( is.null(nextGen) | (min(nextGen$births) > maxTime) ) {
           # If complete after maxTime set complete to NA
      allGens = lapply(allGens, function(gen) {
        gen$completes[ gen$completes > maxTime ] = NA
        gen
      })
      return(allGens)
    }
    
    # Drop those born after maxTime 
    nextGen = nextGen[ nextGen$births <= maxTime , ]
    nextGen$kidID = 1:nrow(nextGen)
    numGens = numGens + 1
    allGens[[ numGens ]] = nextGen    
  }  
}	
```


We may want to rework the familytree function to drop all children born after a certain time limit and also to continue running until an entire generation is born after the time limit.  We could also limit the total number of offspring instead of the time.  We will do that below.
```{r}
familyTree = function(lambda = 0.5, kappa = 0.3, 
                      maxGen = 10, maxOffspring = 1000) {

       # Return value - a list with 1 data frame per generation.
  allGens = vector(mode = "list", length = maxGen)
  
       # Generate root of the tree
  allGens[[1]] = data.frame(parentID = NA, kidID = 1, 
                            births = 0, 
                            completes = rexp(1, rate = kappa))
  
  currentNumOffspring = 0

       # Generate future generations, one at a time.
  for (i in 2:maxGen) {
    nextGen = genKidsV(bTimes = allGens[[ (i - 1) ]]$births,
                       cTimes = allGens[[ (i - 1) ]]$completes,
                       parentID = allGens[[ (i - 1) ]]$kidID,
                       lambda = lambda, kappa = kappa)
    if (is.null(nextGen)) return(allGens[ 1:(i - 1) ]) 
    allGens[[ i ]] = nextGen
    currentNumOffspring = currentNumOffspring + nrow(nextGen)
    if (currentNumOffspring > maxOffspring) 
      return(allGens[1:i])
  }  
  allGens
}
```
Let's see what happens when running the new version of familyTree with the same settings as before.
```{r}
set.seed(seed2)
tree = familyTree(lambda = 1, kappa = 0.5, maxGen = 100, maxOffspring = 1000)
length(tree)
```
The tree now only produces 7 generations before hitting the cap of 1000 children.
```{r}
sapply(tree, nrow)
```

# 7.9 - Replicating the Simulation
Now we can write a function to run the familyTree function multiple times and summarize the results.  First we write a helper function to generate the summary statistics.
```{r}
exptOne = function(l, k, mG, mO){
  # Helper function to call familyTree
  # Returns - summary statistics for analysts
  
  aTree = familyTree(lambda = l, kappa = k, maxGen = mG, maxOffspring = mO)
  numGen = length(aTree)
  numJobs = sum(sapply(aTree, nrow))
  c(numGen, numJobs)
}
```
Let's try running this function with the previously used lambda, kappa, maxGen, and maxOffspring settings.
```{r}
set.seed(seed2)
exptOne(1, 0.5, 100, 1000)
```
Now we can write another function to call the helper function multiple times.
```{r}
MCBA = function(params, repeats = 5, mG = 10, mO = 1000){
  # params: matrix columns of lambda and kappa values
  # For each lambda and kappa pair, run "repeats" times
  
  
  n = nrow(params)
  
  mcResults = vector("list", length = n)
  
  for (i in 1:n) {
    cat("param set is ", i, "\n")
    mcResults[[i]] = replicate(repeats,
                               exptOne(l = params[i, 1],
                                       k = params[i, 2],
                                       mG = mG, mO = mO))
  }
  mcResults
}
```
Now we can test out this function
```{r}
trialKappas = c(0.1, 10, 0.1, 10)
trialLambdas = c(0.1, 0.1, 10, 10)
trialParams = matrix(c(trialLambdas, trialKappas), ncol = 2)
mcTrialOutput = MCBA(params = trialParams, repeats = 100,
                     mG = 200, mO = 100000)
```

